Concurrent Programming with CXL

### 2020-05-21 book/paper.tex
Starting on a "let" expression to replace "var" eventually
Latest

Again, square brackets and parentheses work the same in CXL, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same CXL value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same CXL value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if \texttt{x == [ False, True ]},
then \texttt{x[0] == False} and \texttt{x[1] == True}, just like in Python.

@@

The final na\"{\i}ve solution is based on a variable called \texttt{turn}
that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
enter the critical section, while process $1-i$ has to wait.  When done,
process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
opportunity to enter.
An invariant of this solution is that while process $i$ is in the critical
section, \texttt{turn} == $i$.
(If you do not believe that, replace the assertion at the critical
section with \texttt{assert turn == self} and run CXL.)
Since \texttt{turn} cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that
processes 0 and 1 alternate entering the critical section.
The problem with the solution is
that it also violates the \emph{progress} property:
if process~$i$ terminates instead of entering the critical section when it
is process~$i$'s turn, process $1-i$ ends up waiting indefinitely for its
turn.

@@

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of $N$ processes.
Variable \texttt{shared} is initialized to \texttt{False} while
\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
An important invariant, $I_1$, of the program is that at any time at most
one of the variables is \texttt{False}.
Another invariant, $I_2$, is that if process $i$ is in the critical section,
then \texttt{private[$i$] == False}.
Between the two, it is clear that only one process can be in the
critical section at the same time.

@@

Invariant $I_2$ is obvious as process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False} and no other process modified
\texttt{private[$i$]}.
Indeed, there is no need to keep \texttt{private} as a shared
variable---we only did so to be able to show the invariants.
Figure~\ref{fig:tas2} shows a more straightforward implementation.
Compared to Peterson's Algorithm, this approach is general for
any number of processes.

@@

In CXL, atomic statements allow you to implement your own
low synchronization primites like test-and-set.  Atomic statements
are not intended to replace locks or other synchonization primitives.
By the way, CXL \texttt{assert} statements execute atomically
as if they are nested within an atomic statement.

>>>>

Again, square brackets and parentheses work the same in CXL, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same CXL value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same CXL value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if \texttt{x == [ False, True ]},
then \texttt{x[0] == False} and \texttt{x[1] == True}, just like in Python.
However, when creating a singleton list, make sure you include the
comma, as in \texttt{[False,]}.

@@

The final na\"{\i}ve solution is based on a variable called \texttt{turn}
that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
enter the critical section, while process $1-i$ has to wait.  When done,
process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
opportunity to enter.
An invariant of this solution is that while process~$i$ is in the critical
section, \texttt{turn} == $i$.
(If you do not believe that, replace the assertion at the critical
section with \texttt{assert turn == self} and run CXL.)
Since \texttt{turn} cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that
processes 0 and 1 alternate entering the critical section.
The problem with the solution is
that it also violates the \emph{progress} property:
if process~$i$ terminates instead of entering the critical section when it
is process~$i$'s turn, process $1-i$ ends up waiting indefinitely for its
turn.

@@

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of $N$ processes.
It uses $N+1$ variables.
Variable \texttt{shared} is initialized to \texttt{False} while
\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
An important invariant, $I_1$, of the program is that at any time at most
one of these variables is \texttt{False}.
Another invariant, $I_2$, is that if process $i$ is in the critical section,
then \texttt{private[$i$] == False}.
Between the two, it is clear that only one process can be in the
critical section at the same time.

@@

Invariant $I_2$ is obvious as process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False} and no other process modifies
\texttt{private[$i$]}.
Indeed, there is no need to keep \texttt{private} as a shared
variable---we only did so to be able to show the invariants.
Figure~\ref{fig:tas2} shows a more straightforward implementation.
Compared to Peterson's Algorithm, this approach is general for
any number of processes.

@@

In CXL, atomic statements allow you to \emph{implement} your own
low synchronization primitives like test-and-set.  Atomic statements
are not intended to \emph{replace} locks or other synchonization primitives.
By the way, CXL \texttt{assert} statements execute atomically
as if they are nested within an atomic statement.
